---
title: "Getting started with islhr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with islhr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4
)
library(islhr)
islh_setup(format = "html", quiet = TRUE)
```

`islhr` gives figures, tables and Quarto reports one consistent Island Health
visual system. This guide starts with working output, then explains which
function to use for each task.

## Choose a starting point

| I want to... | Start with | What it returns |
|---|---|---|
| Configure a report | `islh_setup()` | Plot and table defaults for the current format |
| Make a branded plot | `theme_islh()` and a `scale_*_islh()` function | A ggplot object |
| Make an HTML table | `islh_gt()` | A `gt` table |
| Make a Word table | `islh_flextable()` | A `flextable` object |
| Start a report project | `islh_create_report()` | A ready-to-render Quarto project |
| Find a colour | `islh_brand()` or `islh_hex()` | Hex colour values |
| Draw a health-geography map | `theme_islh_map()` and `coord_islh_map()` | Map theme and coordinates |

## A complete first example

The package includes a small data set so every example can run without a
network connection.

```{r example-data}
counts <- islh_example_data()
counts
```

An ordinary ggplot becomes an Island Health figure by adding the appropriate
scale. `islh_setup()` has already applied the theme.

```{r example-figure, fig.alt="Bar chart showing encounters for three example Island Health programmes."}
library(ggplot2)

encounter_plot <- ggplot(counts, aes(program, encounters, fill = program)) +
  geom_col(show.legend = FALSE) +
  scale_fill_islh() +
  scale_y_islh_count() +
  labs(
    title = "Encounters by programme",
    subtitle = "Example data included with islhr",
    x = NULL,
    y = "Encounters"
  )

encounter_plot
```

For HTML, pass the same data frame to `islh_gt()`. The returned object is still
a regular `gt` table, so formatting can be added before or after applying the
theme.

```{r example-table}
islh_gt(
  counts,
  title = "Programme activity",
  subtitle = "Illustrative values",
  source_note = "Source: simulated package data."
)
```

For Word, replace that call with `islh_flextable(counts)`. The Word table is
rendered by Word rather than the browser, so it is demonstrated in the
generated report project instead of duplicated on this HTML page.

## Set up a report

Put one setup call near the top of a Quarto document:

````
```{r}
#| label: setup
#| include: false
library(islhr)
islh_setup()
```
````

With `format = "auto"`, the default, `islh_setup()` detects HTML or Word and
selects the matching table system. Outside a render it configures plots only.
Loading the package by itself does not change themes or options.

Check optional dependencies before rendering a report:

```{r dependency-check}
check <- islh_check("html", quiet = TRUE)
check[c("format", "tables", "ok")]
```

If `ok` is `FALSE`, run `islh_install_deps("html")`, `"docx"` or `"both"` once
on that computer.

## Colours and scales

Use role names when the colour has a meaning. This keeps the report aligned if
the brand tokens change later.

```{r brand-colours}
islh_brand("primary")
islh_brand(c("success", "warning", "danger"))
islh_hex("blue", c(30, 50, 80))
```

Use the scale that matches the data:

| Data meaning | Function | Typical use |
|---|---|---|
| Unordered categories | `scale_colour_islh()` or `scale_fill_islh()` | Programmes or regions |
| Ordered categories | `scale_*_islh_ordinal()` | Low, medium and high |
| Status | `scale_*_islh_signal()` | Good, warning and bad |
| Continuous values in bands | `scale_fill_islh_b()` | Choropleth maps |

The qualitative scale is intentionally limited to five categories. If the
plot has more groups, combine small groups, facet the figure or label lines
directly instead of asking colour alone to carry the distinction.

## Save output consistently

`islh_save_plot()` applies standard report or slide dimensions and uses a
high-quality raster device when available.

```{r save-plot, eval=FALSE}
islh_save_plot("encounters.png", encounter_plot, preset = "report")
```

The function writes a file, so the call is not run while this website builds.
It returns the saved path invisibly.

## Start a full report

```{r create-report, eval=FALSE}
islh_create_report("2026-flu-season", format = "both")
```

The project contains a working report, the Island Health Quarto extension,
brand configuration and example data. Open the generated `.Rproj`, replace the
example values, and render. See [Writing reports](reports.html) for the complete
workflow and [Mapping BC health geographies](maps.html) for maps.

## Common problems

- A table function reports a missing package: run `islh_check()` and install
  the dependency set for the output format.
- A plot uses an unexpected font: run `islh_font_family()` to see which family
  was selected. BC Sans is not bundled with the package.
- A Word render cannot replace its output: close the `.docx` file before
  rendering again.
- A map looks stretched: add `coord_islh_map()` so it uses BC Albers.

Run `islh_help()` for a compact console reference when working interactively.
