---
title: "Getting started with islhr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with islhr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`islhr` puts Island Health branding on the things a report is made of:
figures, tables, and the page they sit on. This article covers the parts you
will use in the first hour.

## Install what your format needs

`islhr` itself has few dependencies. The packages that draw tables are
different for HTML and Word, and you only need the ones for the format you are
writing, so they are optional and installed on request:

```r
islh_install_deps("docx")   # or "html"
```

`islh_check()` tells you where you stand without installing anything:

```{r}
library(islhr)
islh_check("plots", quiet = TRUE)$required
```

It checks versions as well as presence. Two are worth knowing about because
the failure without them is confusing: the colour scales need ggplot2 3.5.0 or
newer, and `islh_flextable()` needs flextable 0.9.10 or newer.

## Set up a report

One call at the top of a Quarto document:

```r
library(islhr)
islh_setup()
```

`islh_setup()` looks at what you are rendering to and configures the rest: the
ggplot2 theme and default colours, the matching table engine (`gt` for HTML,
`flextable` for Word), and the BC Sans font if it is installed. Outside a
render it configures plots only.

Loading the package does none of this. `library(islhr)` only defines
functions — it sets no options, changes no theme and prints nothing — so
attaching it can never change how an unrelated plot looks.

## Colours

Use the role names rather than hex values, so a change to the brand reaches
your report without you editing it:

```{r}
islh_brand("primary")
islh_brand(c("success", "warning", "danger"))
```

The full ramp is there when you need a specific step. Each of the eight
families has twenty steps, running from 10 (dark) to 98 (light). Ask for a
step that does not exist and the error lists the ones that do.

```{r}
islh_hex("blue", c(30, 50, 80))
islh_hex("grey", 50)
```

## Figures

`islh_setup()` sets the theme globally, so an ordinary ggplot is already
branded. The scales are there when you are mapping a variable to colour:

```r
ggplot2::ggplot(data, ggplot2::aes(month, cases, colour = programme)) +
  ggplot2::geom_line() +
  scale_colour_islh() +
  scale_y_islh_count()
```

`scale_colour_islh()` is capped at five categories by design — beyond that the
colours stop being reliably distinguishable. Use `scale_colour_islh_ordinal()`
for ordered categories and `scale_*_islh_signal()` for good/warning/bad status,
always paired with a label or shape so the meaning does not depend on colour
alone.

## Tables

```r
islh_gt(counts)          # HTML
islh_flextable(counts)   # Word
```

Both fill the text width. Pass `width = 0.6` for something narrower.

## Small counts

```{r}
islh_suppress(c(0, 3, 42), threshold = 5)
```

There is deliberately no default threshold. The right rule depends on the data
and on the release, and a default would make that decision silently.

For a whole table, and for the case where a visible total lets a reader
recover a suppressed cell by subtraction:

```{r}
counts <- data.frame(area = c("North", "Central", "South"), cases = c(3, 42, 17))
islh_suppress_table(counts, "cases", threshold = 5, complementary = TRUE)
```

## Where to go next

`islh_help()` prints a grouped list of the functions you will use most.
`vignette("reports", package = "islhr")` walks through a full report.
