---
title: "Mapping BC health geographies"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mapping BC health geographies}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

The two packages divide mapping work by responsibility. `islhepi` retrieves
and standardizes population and health-geography data. `islhr` controls the
visual presentation of the finished map.

## Retrieve and join the data

The following example requests total population and LHA boundaries for Island
Health. The catalogue calls are not evaluated while this vignette is built.

```{r, eval=FALSE}
library(dplyr)
library(ggplot2)
library(islhepi)
library(islhr)

population_extract <- islh_bc_population(
  geography = "lha",
  years = 2025,
  sex = "T"
)

population <- population_extract |>
  summarise(
    population = sum(population),
    .by = c(geography, geography_code, year, estimate_type)
  )

boundaries <- islh_bc_geography("lha")

map_data <- left_join(
  boundaries,
  population,
  by = join_by(geography, geography_code),
  relationship = "one-to-one",
  na_matches = "never"
)

if (anyNA(map_data$population)) {
  stop("One or more Island Health boundaries did not match population data.")
}
```

The population resource covers all of BC, while the default boundary object
contains only Island Health. Extra provincial population rows are expected to
be discarded. The one-to-one relationship rejects duplicate codes, and the row
check catches any Island Health boundary without a population match.

## Draw the choropleth

`theme_islh_map()` keeps titles, captions and legends from `theme_islh()` but
removes axes, ticks and grids. `scale_fill_islh_b()` places a continuous value
into branded colour steps.

```{r, eval=FALSE}
ggplot(map_data, aes(fill = population)) +
  geom_sf(colour = islh_hex("grey", 80), linewidth = 0.25) +
  scale_fill_islh_b(labels = scales::label_comma()) +
  labs(
    title = "Population by Local Health Area",
    subtitle = "Island Health, 2025",
    fill = "Population",
    caption = "Source: BC Stats and BC Data Catalogue"
  ) +
  theme_islh_map()
```

For a categorical map, use `scale_fill_islh()` instead. A map should retain a
text label or table alternative when colour is the only visual distinction.

## Reproducibility

Capture `attr(population_extract, "bcdata_source")` and the corresponding
attribute on `boundaries` before transforming them. Save a dated extract in
the analysis project when a report must reproduce the exact population or
boundary version used for publication.
