---
title: "Mapping BC health geographies"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mapping BC health geographies}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.8
)
library(ggplot2)
library(islhr)
# Resolve the brand font first, as the other articles do. Unresolved, the
# theme names BC Sans, and where it is not installed the PDF device that
# R CMD check runs this code on stops with "invalid font type".
islh_setup(format = "plots", quiet = TRUE)
```

The two packages divide mapping work by responsibility. `islhepi` retrieves
and standardizes population and health-geography data. `islhr` controls the
visual presentation of the finished map.

## See the map functions work

This first example uses four simple polygons created in the vignette. It does
not contact the BC Data Catalogue, so the website can display a real map while
remaining reproducible.

```{r map-example-data}
demo_map <- data.frame(
  geography_name = c("North", "Central", "West", "South"),
  population = c(12000, 34000, 18000, 52000),
  wkt = c(
    "POLYGON ((0 1, 1 1, 1 2, 0 2, 0 1))",
    "POLYGON ((1 1, 2 1, 2 2, 1 2, 1 1))",
    "POLYGON ((0 0, 1 0, 1 1, 0 1, 0 0))",
    "POLYGON ((1 0, 2 0, 2 1, 1 1, 1 0))"
  )
)
demo_map <- sf::st_as_sf(demo_map, wkt = "wkt", crs = 3005)
demo_map[c("geography_name", "population")]
```

```{r map-example, fig.alt="Four-area demonstration choropleth, with the southern area darkest because it has the largest illustrative population."}
ggplot(demo_map, aes(fill = population)) +
  geom_sf(colour = islh_hex("grey", 80), linewidth = 0.4) +
  scale_fill_islh_b(
    n.breaks = 4
  ) +
  coord_islh_map() +
  labs(
    title = "Population by area",
    subtitle = "Demonstration geometry and values",
    fill = "Population",
    caption = islh_caption(
      source = "Simulated",
      extracted = "2026-09-04"
    )
  ) +
  theme_islh_map()
```

The same three presentation functions work with downloaded LHA, HSDA, CHSA or
health-authority boundaries. The next section replaces `demo_map` with real
catalogue data.

## Retrieve and join the data

The following example requests total population and LHA boundaries for Island
Health. Catalogue calls are shown but not evaluated during the website build,
because the upstream service can be unavailable even when the package is
working correctly.

```{r, eval=FALSE}
library(dplyr)
library(ggplot2)
library(islhepi)
library(islhr)

population_extract <- islh_bc_population(
  geography = "lha",
  years = 2025,
  sex = "T"
)

population <- population_extract |>
  summarise(
    population = sum(population),
    .by = c(geography, geography_code, year, estimate_type)
  )

boundaries <- islh_bc_geography("lha")

map_data <- left_join(
  boundaries,
  population,
  by = join_by(geography, geography_code),
  relationship = "one-to-one",
  na_matches = "never"
)

if (anyNA(map_data$population)) {
  stop("One or more Island Health boundaries did not match population data.")
}
```

The population resource covers all of BC, while the default boundary object
contains only Island Health. Extra provincial population rows are expected to
be discarded. The one-to-one relationship rejects duplicate codes, and the row
check catches any Island Health boundary without a population match.

## Clip the boundaries to land

Health boundaries extend over water. Drawn unclipped, Island Health becomes a
blob across the Salish Sea. Clip to a terrestrial layer first, then simplify:

```{r, eval=FALSE}
map_data <- map_data |>
  sf::st_intersection(bc_land) |>
  rmapshaper::ms_simplify(keep = 0.05, keep_shapes = TRUE)
```

Use `rmapshaper::ms_simplify()` rather than `sf::st_simplify()`. It preserves
shared borders, so adjacent areas do not end up with slivers between them, and
`keep_shapes = TRUE` keeps the small islands. Neither package is a dependency
of `islhr`; install them in the analysis project.

## Draw the choropleth

Three functions carry the presentation:

* `theme_islh_map()` starts from `ggplot2::theme_void()` and adds the brand
  text hierarchy, so nothing but the data draws.
* `coord_islh_map()` locks the map to BC Albers (EPSG:3005) and removes the
  graticule.
* `scale_fill_islh_b()` places a continuous value into branded colour steps.

```{r, eval=FALSE}
ggplot(map_data, aes(fill = population)) +
  geom_sf(colour = islh_hex("grey", 80), linewidth = 0.25) +
  scale_fill_islh_b() +
  coord_islh_map() +
  labs(
    title = "Population by Local Health Area",
    subtitle = "Island Health, 2025",
    fill = "Population",
    caption = islh_caption(
      source = "BC Stats and BC Data Catalogue",
      extracted = "2026-03-31",
      boundary = "Local health areas, 2024 boundaries"
    )
  ) +
  theme_islh_map()
```

Always set the projection. Latitude and longitude (EPSG:4326) stretches
Vancouver Island sideways and misrepresents how large each area is. BC Albers
is what the provincial layers already use.

The legend abbreviates large counts, so `25000` prints as `25K`. A bottom
legend has little room between breaks, and full comma-separated numbers
overlap once a map has more than three or four bins. Pass
`labels = scales::label_comma()` when a figure needs exact numbers, along with
`key_width = 16` to lengthen the bar.

For a categorical map, use `scale_fill_islh()` instead. A map should retain a
text label or table alternative when colour is the only visual distinction.

## Put the legend in the ocean

Island Health runs northwest to southeast, which leaves open water off the
west coast and empty mainland in the top right. `coord_sf()` fixes the aspect
ratio, so a wider figure cannot stretch the map to fill the space a bottom
legend takes. Moving the legend inside the panel gives the map roughly a fifth
more room:

```{r, eval=FALSE}
theme_islh_map(legend = "inside")
```

Adjust `legend_inside` if the default position covers part of the coast. It
takes fractions of the panel measured from the bottom left.

## Map rates, and classify on purpose

Map age-standardized rates rather than counts, and name the standard
population in the caption. A count map of local health areas mostly shows
where people live.

Choose the bin breaks deliberately. `classInt::classIntervals()` offers
quantile, Jenks and equal-interval methods; quantile suits comparisons between
areas, equal interval suits the same indicator across time. Then hard-code the
breaks when a report maps the same indicator more than once:

```{r, eval=FALSE}
scale_fill_islh_b(breaks = c(50, 100, 150, 200))
```

Bins recalculated from each year's data make an unchanged pattern look like a
change.

Five to seven bins is the ceiling. Beyond that nobody can match a swatch back
to the map.

## Keep suppressed, zero and missing apart

Suppressed, zero and no data are three different statements and must read as
three different things.

`scale_fill_islh_b()` fills areas with no data in a grey lighter than every
bin, so a greyscale print still separates them from the lowest band. Draw
suppressed areas as their own layer with their own key:

```{r, eval=FALSE}
suppressed <- filter(map_data, cases < 5)
released <- filter(map_data, cases >= 5)

ggplot(released, aes(fill = rate)) +
  geom_sf(colour = islh_hex("grey", 80), linewidth = 0.25) +
  scale_fill_islh_b() +
  ggnewscale::new_scale_fill() +
  geom_sf(
    data = suppressed,
    aes(fill = "Suppressed"),
    colour = islh_hex("grey", 60),
    linewidth = 0.25
  ) +
  ggplot2::scale_fill_manual(
    values = c(Suppressed = islh_hex("grey", 60)),
    name = NULL,
    aesthetics = "fill"
  )
```

This example uses `ggnewscale` to give the suppression layer its own legend.
For a hatch rather than a flat fill, `ggpattern::geom_sf_pattern()` does the
same job. Give zero counts their own baseline tint through an explicit lowest
break so a reader does not read zero cases as uncollected data.

## Insets for the urban areas

Mount Waddington and Strathcona dominate the page while Greater Victoria, where
most of the population lives, shrinks to a few small polygons. Build the dense
areas as their own panels and combine them:

```{r, eval=FALSE}
victoria <- base_map +
  coord_islh_map(xlim = c(1180000, 1230000), ylim = c(360000, 400000)) +
  labs(title = NULL, subtitle = NULL, caption = NULL) +
  theme_islh_map(legend = "none")

patchwork::wrap_plots(base_map, victoria, widths = c(2, 1))
```

`patchwork::inset_element()` places the panel over the main map instead. Draw a
rectangle on the overview showing the inset extent either way.

If the report is about people rather than land, consider a tile map or a
population cartogram; a choropleth of rural areas is mostly a map of geography.

## Give each local health area its own colour

Some figures compare areas rather than map a value: which LHA is which on a
reference map, or a line for each LHA over time. `islh_areas()` gives each of
the 14 LHAs a fixed colour, the same in every report, and
`scale_fill_islh_area()` and `scale_colour_islh_area()` apply them.

No palette makes 14 areas identifiable by colour alone, so the colours work in
two levels:

* The health service delivery area picks the colour family: Blue for South,
  Cedar for Central and Thistle for North Vancouver Island. These are the
  brand's low-signal families, meant for organizing content. Red, Green and
  Orange keep their signal meanings.
* The LHA picks the shade within the family. Neighbouring LHAs differ by at
  least 20 values within an HSDA, and South and North never touch, so the
  families that colour-blind readers confuse most never meet on a map.

```{r lha-colours}
islh_areas()
```

The brand standard allows more than two colour families only when colour
carries meaning, and then never as the only cue. So label the areas. On a
chart, one panel per HSDA holds 4 to 6 lines of one family, and a name at the
end of each line does the work a legend would:

```{r lha-lines, fig.height = 7, fig.alt = "Three panels, one per health service delivery area, each showing simulated weekly rates for its local health areas as lines in shades of one colour family, named at the right-hand end."}
weeks <- seq(as.Date("2026-01-04"), by = "week", length.out = 10)
areas <- islh_areas()
rates <- expand.grid(week = weeks, code = areas$code, stringsAsFactors = FALSE)
rates <- merge(rates, areas[c("code", "name", "hsda")])
rates$hsda <- factor(rates$hsda, levels = unique(areas$hsda))
level <- 8 * as.numeric(substr(rates$code, 3, 3))
rates$rate <- 10 + level + 6 * sin(as.numeric(rates$week - min(weeks)) / 20)
last_week <- rates[rates$week == max(weeks), ]

ggplot(rates, aes(week, rate, colour = name)) +
  geom_line(linewidth = 0.9) +
  geom_text(
    data = last_week,
    aes(label = name),
    colour = islh_hex("grey", 25),
    hjust = 0,
    nudge_x = 2,
    size = 3
  ) +
  facet_wrap(vars(hsda), ncol = 1) +
  scale_x_date(expand = expansion(mult = c(0.02, 0.45))) +
  scale_colour_islh_area(guide = "none") +
  labs(
    title = "Weekly rate by local health area",
    subtitle = "Simulated values",
    x = NULL,
    y = "Rate per 100,000"
  ) +
  theme_islh()
```

Label text sits beside the line in dark grey, not in the line's colour: small
text needs 70 values of contrast with the page, which the lighter shades do
not have.

On a map, write the code or name on each area in its `label_colour`, which is
white or Grey 10, whichever the fill needs. That contrast holds for large
labels only, about 14 pt and bold. Put smaller labels outside the shapes, and
give the small South Island areas an inset, as above.

```{r, eval=FALSE}
lha <- merge(islh_bc_geography("lha"), islh_areas(), by.x = "geography_code",
             by.y = "code")

ggplot(lha) +
  geom_sf(aes(fill = name), colour = "white", linewidth = 0.3) +
  geom_sf_text(
    aes(label = geography_code, colour = I(label_colour)),
    size = 5,
    fontface = "bold"
  ) +
  scale_fill_islh_area(name = NULL) +
  coord_islh_map() +
  labs(title = "Island Health local health areas") +
  theme_islh_map(legend = "none")
```

The scales match the official names or the three-digit codes, so data from
`islh_bc_geography()` works as it comes. A value that matches no LHA, such as
a misspelled name, is drawn grey with a warning naming it.
`scale_fill_islh_area("hsda")` does the same for the three HSDAs, in each
family's representative colour.

## Colour and contrast

The map palette is a single-hue blue ramp whose steps fall about ten points of
CIE lightness apart. Single hue plus even lightness steps is what keeps a ramp
readable in greyscale and for colour-blind readers, so the brand colours do not
need replacing with Viridis here. Check any palette you build yourself with
`colorspace::specplot()`, view the finished map through `colorspace::deutan()`,
and print it in greyscale once. Internal readers print.

Use a diverging scale only when the map compares against a baseline, such as a
standardized ratio against the Island Health average. "Where is this worse than
expected" is usually the question, and a ratio answers it more directly than a
rate.

## Alt text

Every map chunk needs `fig-alt` in Quarto. Describe what the map shows, not
that it is a map: the geography, the indicator, and the pattern a reader would
see.

````
```r
#| label: fig-lha-rate
#| fig-alt: >
#|   Choropleth of Island Health local health areas showing influenza
#|   hospitalization rates. Rates are highest in the north Island and
#|   lowest in Greater Victoria.
```
````

## Reproducibility

Capture `attr(population_extract, "bcdata_source")` and the corresponding
attribute on `boundaries` before transforming them. Save a dated extract in
the analysis project when a report must reproduce the exact population or
boundary version used for publication.
