---
title: "Moving from the theme script"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Moving from the theme script}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

If your project has an `R/island-health-ggplot-theme.R` in it, it came from a
starter-kit ZIP. This is how to move it to the package.

## What changes

Nothing about the function names. `theme_islh()`, `islh_setup()`,
`islh_flextable()` and the rest work exactly as before. What changes is where
they come from: the package, rather than a copy of the script inside your
project. A fix then reaches your report the next time you update `islhr`,
instead of the next time you download a ZIP.

## The three steps

**1. Install the package.**

```{r}
install.packages("remotes", type = "binary")
remotes::install_github("temuulene/islhr")
```

**2. Change the setup chunk.** Replace:

```{r}
source("R/island-health-ggplot-theme.R")
islh_setup()
```

with:

```{r}
library(islhr)
islh_setup()
```

**3. Delete the copied script**, and the `R/` folder if it holds nothing else.

For a Word report, also run `islh_use_quarto()` and point the front matter at
`format: islh-report-docx`. That replaces the loose `islh-report-reference.docx`
and the hand-written `reference-doc:` line; you can delete the local copy of
the reference document.

## Things that are gone

- **`islh_theme_version`** is replaced by `islh_version()`, which reads the
  installed package version. There is no longer a version string inside a
  script to keep in step with anything.
- **`install-packages.R`** is replaced by `islh_install_deps()`. It follows the
  same rules — base R, binary packages, never `pak` — but lives in one place
  rather than in a copy inside every project.
- **The starter-kit ZIPs** are replaced by `islh_create_report()`.

## One thing that may now fail loudly

The package declares minimum versions the script never did:

- **ggplot2 3.5.0** for the colour scales. On an older ggplot2 they used to
  fail inside ggplot2 with `argument "scale_name" is missing`.
- **flextable 0.9.10** for `islh_flextable()`. On an older flextable it used to
  fail with `unused argument (repeat_headers = TRUE)`.

If a report that used to run now stops with a clear message naming one of
these, that is the version gate doing its job. `islh_install_deps()` will
bring the package up to date.

## If you cannot install packages

Some machines will not let you. `islh_check()` will tell you what is missing,
and your team lead can supply a built copy of `islhr` to install from a file:

```{r}
install.packages("path/to/islhr_0.1.0.zip", repos = NULL)
```
