---
title: "Writing reports"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Writing reports}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Start a project

```{r}
islhr::islh_create_report("flu-season-2026", format = "docx")
```

That writes a Quarto project that renders as it stands. Open the `.Rproj`
file, run `islh_install_deps("docx")` once on that machine, then open
`report.qmd` and click **Render**. Replace the example data with your own.

`format` takes `"html"`, `"docx"` or `"both"`. With `"both"`, the report
chooses its table engine at render time, so one file produces either output.

## Add the format to a project you already have

```{r}
islhr::islh_use_quarto()
islhr::islh_use_brand()
```

The first copies the Island Health Quarto extension into `_extensions/`; the
second copies `_brand.yml` alongside it. Quarto only looks for extensions
inside the project directory, which is why these are copies rather than
references into your R library. Re-run `islh_use_quarto()` after upgrading
`islhr` to pick up format changes.

Then name the format:

```yaml
---
title: "Respiratory season summary"
format: islh-report-docx    # or islh-report-html
---
```

## The setup chunk

Every Island Health report starts the same way:

````
```{r}
#| label: setup
#| include: false
library(islhr)
islh_setup()
```
````

## Figures

Give every figure a caption and alt text. Alt text is what a screen-reader
user gets instead of the picture, so describe what the figure *shows*, not
that it is a bar chart:

````
```{r}
#| label: fig-encounters
#| fig-cap: "Encounters by programme"
#| fig-alt: "Primary care had the most encounters at 326, then mental health
#|   at 184, then public health at 79."
#| fig-width: 6
#| fig-height: 3.2
```
````

Reference it in prose with `@fig-encounters` so the numbering stays right when
you move things around.

## Word output

The Word format uses a reference document built from the official Island
Health letterhead: the logo header, the `Island Health / Page X of Y` footer,
BC Sans, and headings in the letterhead's dark blue.

Two things to know:

- The table of contents renders empty. Word populates it as a field the first
  time the document is opened. That is Word's behaviour, not a fault in the
  template.
- Word locks a file it has open. Close `report.docx` before re-rendering, or
  the render fails on a permission error.

## Troubleshooting

**BC Sans is not being used.** It is not shipped with the package, so it has to
be installed on your machine — once, from
[bcgov/bc-sans](https://github.com/bcgov/bc-sans). `islh_font_family()` reports
what was found. Without it, output falls back to another font; everything else
still works.

**Windows plot pane warns about the font database.** Base graphics devices on
Windows resolve font names through R's own font database rather than through
`systemfonts`. `islh_setup()` registers the family to stop the warnings. If
they persist, set the RStudio graphics backend to AGG under
*Tools > Global Options > General > Graphics*.

**Figures vanish from the rendered document.** On Windows, a path longer than
260 characters silently drops figures. Keep the project somewhere short, not
several folders deep in a synced drive.

**A package is missing or too old.** `islh_check()` names it and gives you the
command. `islh_install_deps()` runs that command for you. Both use base R's
`install.packages()` with `type = "binary"`, because Island Health laptops
block installers that unpack with their own helper binary and have no compiler
for a source build.

## Before anything leaves your team

Reports built this way are ordinary files with no access controls. Never put
identifiable information in one, and suppress small cells before sharing:

```{r}
counts$cases <- islh_suppress(counts$cases, threshold = 5)
```

`islh_suppress_table()` does a whole table, and its `complementary` argument
handles the case where a visible total would let a reader recover a suppressed
cell by subtraction. The threshold is yours to set — check the rule your
release is governed by.
